---
title: "ORA with EnrichmentBrowser, phyper, and fisher.test"
output:
    html_document:
        toc: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, message=FALSE}
library(EnrichmentBrowser)
library(dplyr)
library(purrr)
```

## Create example data

Create example data using the functions in the EnrichmentBrowser package:

```{r}
seed <- 123
sig_gen_fdr <- 0.1 #FDR threshold for significant genes in the DA

## Create SummarizedExperiment
set.seed(seed)
se <- makeExampleData(what="SE")

## Run Differential abundance
set.seed(seed)
se <- deAna(se)

# Create a list of ten signatures
set.seed(seed)
gs <- makeExampleData(what="gs", gnames=names(se))

n_unique_g <- length(unique(unlist(gs, use.names = FALSE)))
```

The SummarizedExperiment (SE) contains `r nrow(se)` unique genes in total:

```{r}
se
```

There are `r n_unique_g` unique genes across all signatures.

```{r}
n_unique_g
```

Individual lengths of the signatures:

```{r}
map_int(gs, length)
```

Define significant genes (adjusted pvalue < 0.1):

```{r}
sig.genes <- names(se)[rowData(se)$ADJ.PVAL < sig_gen_fdr]
length(sig.genes)
```
Two of the signatures (gs1 and gs2) will be composed of significantly DA genes
entirely:

```{r}
gs[[1]] <- sample(sig.genes, length(gs[[1]])) 
gs[[2]] <- sample(sig.genes, length(gs[[2]]))   

map_dbl(gs, ~ round(mean(.x %in% sig.genes), 2))
```

## Run ORA

In this analysis, we are using all of the original 10 signatures:

```{r}
ea.res <- sbea(
    method = "ora", se = se, gs = gs, perm = 0,
    alpha = sig_gen_fdr, # FDR threshold for significant genes in the DA
    padj.method = "fdr"
)
res <- ea.res$res.tbl |> 
    as.data.frame() |> 
    arrange(PVAL)
res |> 
    knitr::kable()
```

## Reproduce "manually" what sbea does

To inspect the results more in detail, let's reproduce the code
implemented in EnrichmentBrowser, but adding information about the number
of genes in the background as well.

EnrichmentBrowser uses the `phyper` function for ORA.

The following snippets of code were taken from the EnrichmentBrowser
package (in Bioconductor):

```{r, warning=FALSE}
myORA <- function(se, gs, sig.genes) {
    ## Get a subset of signatures and genes in the SE for further analysis
    GS.MIN.SIZE <- 5
    GS.MAX.SIZE <- 500

    ## Subset of the original SE.
    ## This depends on the genes available in the signatures database 
    igenes <- intersect(rownames(se), unique(unlist(gs)))
    newSE <- se[igenes,]

    ## Subset of the genes in the signatures
    newGS <- lapply(gs, function(s) s[s %in% igenes])
    lens <- lengths(newGS)
    newGS <- newGS[lens >= GS.MIN.SIZE & lens <= GS.MAX.SIZE]
    f <- file()
    sink(file = f)
    cmat <- safe::getCmatrix(newGS, as.matrix = TRUE)
    sink()
    close(f)
    newSE <- newSE[rownames(cmat),] 

    ## Subset of the original number of significant genes
    isig <- rownames(newSE) %in% sig.genes
    nr.sigs <- sum(isig)

    ## Run ORA with phyper
    sig.cmat <- cmat & isig
    ovlp.sizes <- colSums(sig.cmat)
    gs.sizes <- colSums(cmat) 
    uni.sizes <- nrow(newSE) - gs.sizes 
    gs.ps <- phyper(
        q = ovlp.sizes - 1, # number of significant genes in the signature (minus 1)
        m = gs.sizes, # number of genes in the signature (significant or not)
        n = uni.sizes, # number of genes not in the signature (significant or not)
        k = nr.sigs, # total number of significant genes (in the signature or not)
        lower.tail = FALSE
    ) 
    
    res_tbl <- tibble(
        gene_set_name = colnames(cmat),
        genes_in_set = gs.sizes,
        sig_genes_in_set = ovlp.sizes,
        p_value = gs.ps,
        genes_not_in_set = uni.sizes,
        total_sig_genes = nr.sigs
    ) |> 
        arrange(p_value)
    
    list(
        rest_tbl = res_tbl,
        newSE = newSE,
        newGS = newGS,
        nr.sigs = nr.sigs
    )
}
my_ora <- myORA(se, gs, sig.genes)
my_ora$rest_tbl |> 
    knitr::kable()
```

Note that after the analysis, the total number of genes in the SE is reduced.
This reduction depends on the intersection between the total unique
genes across the whole list/collection/database of signatures and the genes
present in the SE.

Updated number of genes (reduced):

```{r}
my_ora$newSE
```
Updated number of significant genes (reduced):

```{r}
my_ora$nr.sigs
```

## Check one signature with different accompanying sets of signatures

So, if the number of unique genes in the collection/database of signatures
changes, the background will change as well even when the SE 
and an a target signature are the same:

```{r}
output <- vector("list", length(gs))
for (i in seq_along(output)) {
    ora <- sbea(
        method="ora", se=se, gs=gs[1:i], perm=0, alpha = 0.1, padj.method = "fdr"
    )
    output[[i]] <- as.data.frame(ora$res.tbl)
    names(output) <- paste0("nsigs_", length(1:i))
}
output <- bind_rows(output, .id = "n_gs")
output |> 
    filter(GENE.SET == "gs1") |> 
    knitr::kable()
```


In the example above, the input SE and the signature of interest (gs1) were the
same, but the p-value changed depending on the number of unique genes present
across all signatures.

Total number of genes left in the SE:

```{r}
output2 <- vector("list", length(gs))
for (i in seq_along(output2)) {
    output2[[i]] <- myORA(se, gs[1:i], sig.genes)
    names(output2)[i] <- paste0("nsigs_", length(1:i))
}
map_int(output2, ~ nrow(.x$newSE))
```

Let's use our custom function created above to reproduce the results using
the EnrichemntBrowser functions, but now adding the number of genes in the
background:

```{r, message=FALSE}
res_tbl2 <- output2 |> 
    map(~ return(.x[["rest_tbl"]])) |> 
    bind_rows(.id = "n_gs")
res_tbl2 |>
    filter(gene_set_name == "gs1") |>
    knitr::kable()
```


In order to keep the same p-value regardless of how many signatures
are analyzed, we sould need to add a background as an additional signature.

Let's add a signature for background:

```{r}
background_sig <- list(bk = unique(unlist(gs)))
length(background_sig[[1]])
```

Running the ORA:

```{r}
output3 <- vector("list", length(gs))
for (i in seq_along(output3)) {
    input_gs <- c(gs[1:i], background_sig)
    ora_3 <- sbea(
        method = "ora", se = se, gs = input_gs, perm = 0, alpha = 0.1,
        padj.method = "fdr"
    )
    output3[[i]] <- as.data.frame(ora_3$res.tbl)
    names(output3) <- paste0("nsigs_", length(1:i))
}
output3 <- bind_rows(output3, .id = "n_gs")
output3 |> 
    filter(GENE.SET == "gs1") |> 
    knitr::kable()
```

Above, the pvalue was the same in all cases. FDR will still vary depending
on the number and values of the p-values being adjusted.

## Reproduce with Fisher exact test

Now, let's just create a custom function to achieve the same results, but
using the `fisher.test` function instead of the `phyper` function:

```{r}
myFisher <- function(x) {
    mySE <- x$newSE
    mySigs <- x$newGS
    all_genes <- rownames(mySE)
    results <- vector("list", length(mySigs)) 
    for (i in seq_along(results)) {
        a <- sum(mySigs[[i]] %in% sig.genes)
        b <- sum(!mySigs[[i]] %in% sig.genes)
        c <- sum((all_genes %in% sig.genes) & (!all_genes %in% mySigs[[i]]))
        d <- sum((!all_genes %in% sig.genes) & (!all_genes %in% mySigs[[i]]))
        ct <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)
        pval <- fisher.test(ct, alternative = "g")$p.value
        sig_name <- names(mySigs)[i]
        
        dat <- as.data.frame(matrix(c(sig_name,pval, a, b, c, d), nrow = 1))
        colnames(dat) <- c(
            "sig_name", "p_value",
            "significat_inSignature", "notSignificant_inSignature", 
            "significant_inBackground", "notSignificant_inBackground"
        )
        results[[i]] <- dat
    }
    results |> 
        bind_rows() |> 
        arrange(p_value)
}
```

Reproducing the first analysis:

```{r}
myFisher(output2[[10]]) |> 
    knitr::kable()
```

Reproducing the analysis above with different number of
signatures:

```{r}
output2 |> 
    map(myFisher) |> 
    bind_rows(.id = "n_sigs") |> 
    filter(sig_name == "gs1") |> 
    knitr::kable()
```

## Session information

```{r}
sessioninfo::session_info()
```
